Dissecting the interaction between skill level and mental templates in Late Acheulean handaxe morphology: Archaeological and experimental insights

### 2022-11-24 manuscript/manuscript.Rmd
finished addressing R1's comments.

The morphological variability of Acheulean handaxes has been one of the most well-studied and well-published topics in paleolithic archaeology [@white1998; @earlyhu1998; @key2019]. Despite the recurrent narrative emphasizing the homogeneity and longevity of handaxe assemblages on a global scale and the conservatism behind this phenomenon that evokes genetic explanations [@corbey2016; @corbey2020; @richerson2005; @sterelny2004], many researchers have recognized the diversity within what has been deemed as a unified Acheulean "tradition" and tried to dissect the sources and meaning of this variation [@lycett2008; @nowell2010; @nowell2002; @sharon2011]. More specifically, a complex suite of interconnecting factors have been identified to contribute to handaxe morphological variation, including but not limited to raw material variability [@sharon2008; @eren2014; @lycett2016; @mcnabb2015], percussor properties [@shipton2009], functional differences [@machin2007; @key2016; @kohn1999; @white2018; @key2017], reduction method/intensity [@shipton2009; @shipton2015], learning processes [@lycett2016; @kempe2012], time budgets [@schillinger2014], knapper skill levels [@stout2014; @herzlinger2017; @caruana2021], and mental templates [@garcía-medrano2019; @hutchence2021]. From this extensive list, knapper skill levels and mental templates have been repeatedly mentioned and discussed in the now extensive corpus of handaxe studies, and Boxgrove handaxes have been one of the most studied assemblages from these two angles. Of particular attention here are the experimental works conducted by Stout et al. [-@stout2014] focusing on inferring high knapping skill level and Garcia-Medrano et al. [-@garcía-medrano2019] identifying the mental template of the Boxgrove assemblage. Our paper combines these two perspectives and provides novel insights to the same archaeological assemblage by comparing it with experimentally made handaxes.

In its classical definition, the term mental template indicates that the "idea of the proper form of an object exists in the mind of the maker, and when this idea is expressed in tangible form in raw material, an artifact results" [@deetz1967: 45]. This concept lies at the very foundation of the cultural-historical approach in that the identification of archaeological cultures is based on the existence of distinct mental templates in a given spatial-temporal framework. Early researchers, whether explicitly or implicitly, often endorsed this conceptual framework and actively applied it in the typological analysis of handaxes at the regional level [@roe1969; @wenban-smith2000; @wenban-smith2004]. Combined with the production of large flakes, the emergence of mental templates (or "imposed form") has been recognized as a major technological innovation of the Acheulean compared with the Oldowan [@isaac1986]. For a decade or so, this concept has been less frequently used, since it was criticized for a) its normative and static assumption [@lyman2004], and b) ignorance of other competing factors such as raw material constraints [@white1995]. To avoid the historical baggage associated with this controversial term, some researchers developed alternative frameworks such as "design imperatives" derived from utilitarian and ergonomic principles, which refers to a set of minimum features shared by all handaxes including their glob-butt, forward extension, support for the working edge, lateral extension, thickness adjustment, and skewness [@gowlett2006; @wynn2018].

Until recently, researchers have actively addressed the above-mentioned critiques and reconceptualized the concept of mental template in the study of handaxe morphology. Regarding the normative and static assumptions, Hutchence and Scott [-@hutchence2021], for example, leveraged the theory of "community of practice" [@wenger1998] to explain the stability of Boxgrove handaxe design across multiple generations, especially how the social norms behind the consolidated material expressions were developed and negotiated by individuals in a group who have a shared history of learning. They further emphasized that emergent actions of individual knappers also contribute greatly to the shape of Boxgrove handaxes but they were simultaneously constrained by the imposition of social norms. This view also somewhat echoes the "individualized memic construct" proposed by McNabb et al. [-@mcnabb2004], which highlighted the influence of individual agency that is complementary to the traditionally favored explanation of social learning. As for the critique towards confounding factors explaining morphological variability, raw material is often treated as an important variable to be controlled at the very beginning of a research design focusing on mental templates. This is best exemplified by an experimental study of García-Medrano et al. [-@garcía-medrano2019], where they carefully chose experimental nodules mirroring those found in the Boxgrove archaeological assemblage in composition, size, and shape. In short, when exercised with proper caution, the concept of mental templates still has its value in our study of handaxe morphological variation, which can be further dissected into a series of shape variables corresponding to pointedness, elongation, and cross-sectional thinning among other things.

@@

The negative correlation between the PC1 and PC2 values revealed a hidden structural constraint regarding the relationship between cross-sectional thinning and the imposed form. Our results (Fig.) suggested thinner handaxes (low PC1 value) are generally more pointed/less ovate (high PC2 value). In the thinning phase of handaxe making, a knapper must strike flakes that travel more than one half way across the surface. Consequently, it would be easier to perform thinning if the plan shape of a handaxe is narrower and more pointed. It is possible that such constraints help to explain the convergence of our novice knappers on similar shapes to those prefered by modern expert knappers, however, this clearly does not explain the design target at Boxgrove. Given the ovate forms of the Boxgrove assemblage, it thus requires a high skill level to overcome this structural constraint to produce thin yet wide handaxes as demonstrated by the Boxgrove knappers. This also provides an alternative explanation to the social transmission of form for the experimental convergence of on pointed forms. In this comparative context, it would only be the Boxgrove assemblage that provided evidence of social conformity on a more difficult target shape.

In terms of our second research question, this study shows that training does have an immediate intervention effect (pre-training vs. post-training) in both PC1 (skill level) and PC2 (mental template). Nonetheless, once the training has been initiated, its effects across different assessments on both dimensions are rather non-significant. This finding corroborates what has been suggested in Pargeter et al. [-@pargeter2019] that 90 hours of training for handaxe making is still not enough for novices to reach the skill level as reflected in expert knappers, even considering the massive social support involved in the experiment set up including the direct and deliberate pedagogy and the simplified raw material procurement and preparation procedures. This follow-up project further adds the samples produced by the Late Acheulean toolmaker as a new benchmark to deepen our understanding of this issue. Corroborating Key's [-@keyHandaxeShapeVariation2019] previous finding regarding Boxgrove, it is noteworthy how constrained the range of Boxgrove assemblage morphological variation is as measured by both PC1 and PC2 even when compared with the modern expert group (**Figure** \@ref(fig:GeneralPCA1)), especially given the fact that it has the largest sample size among all studied groups. Some potential explanations for this phenomenon include 1) the strong idiosyncrasy of individual expert knappers shaped by their own unique learning and practice experience; and/or 2) the present day-skill shortage of our expert knapper as compared with Boxgrove knappers despite their multiple years of knapping practice [@milks2019].

@@

Another contribution that we would like to highlight here is that this research project demonstrates the potential of reusing old archaeological data in digital format to address novel research questions. In this paper, the main source of archaeological data is a collection of photos produced and curated more than 10 years ago, and the morphological variation data of the experimental collection are also derived from photographs instead of remeasurements of the original artifacts. Given the irreversible nature of archaeological excavations, digitized data, be it text, pictures, or videos, often become the sole evidence that is available for certain research questions. Yet, it has been widely acknowledged that the reuse of archaeological data has not received enough attention among researchers in our discipline [@faniel2018; @huggett2018; @moody2021]. Among many reasons preventing archaeologists from reusing published and digitized data [@sobotkova2018], the lack of a standardized practice of and motivation for data sharing is a prominent one [@marwick2018]. As stated in the method section, we addressed this issue by sharing the raw data and the code for generating the derived data on an open-access repository. Another major and legitimate concern of archaeological data reuse is their quality. In terms of this aspect, we do acknowledge the limitations of relying on photos when it comes to the more detailed technological analysis of stone artifacts, however, our paper shows that finding the appropriate research questions given the data available is key to revealing new novel insights into the studied topic. Moreover, we believe that this type of research has a strong contemporary relevance due to the continued influence of the COVID-19 on fieldwork-related travel and direct access to archaeological artifacts [@balandier2022; @ogundiran2021].

@@

Regarding the two research questions we proposed in the beginning, our case study suggested that 1) we can delineate the effects of skill level and mental template through the multivariate analysis of morphometric data, where the former is associated with cross-sectional thinning while the latter is reflected in elongation and pointedness; 2) Training has an immediate effect of convergence on shared design targets, but 90 hours of training is still not enough for novice to reach the level of expertise as reflected in modern experienced knappers, let alone the Boxgrove tool makers. At a larger theoretical level it questions the distinction between social learning of design targets vs. individual learning of the skills needed to achieve them. To illustrate, a thin cross section could be part of a mental template or design target and was explicitly instructed by our expert instructor to novices, but novices cannot fully understand nor achieve this technological goal due to the constraint of skill level, making it a robust indicator of the latter. In the future, more robust experimental studies are needed to deepen our understanding of the relationship between skill acquisition and the morphological variability of handaxes as well as their implications for the biological and cultural evolution of the hominin lineages.

>>>>

The morphological variability of Acheulean handaxes has been one of the most well-studied and well-published topics in paleolithic archaeology [@white1998; @earlyhu1998; @key2019]. Despite the recurrent narrative emphasizing the homogeneity and longevity of handaxe assemblages on a global scale and the conservatism behind this phenomenon that evokes genetic explanations [@corbey2016; @corbey2020; @richerson2005; @sterelny2004], many researchers have recognized the diversity within what has been deemed as a unified Acheulean "tradition" and tried to dissect the sources and meaning of this variation [@lycett2008; @nowell2010; @nowell2002; @sharon2011]. More specifically, a complex suite of interconnecting factors [@lycett2015] have been identified to contribute to handaxe morphological variation, including but not limited to raw material variability [@sharon2008; @eren2014; @lycett2016; @mcnabb2015], percussor properties [@shipton2009], functional differences [@machin2007; @key2016; @kohn1999; @white2018; @key2017], reduction method/intensity [@shipton2009; @shipton2015], learning processes [@lycett2016; @kempe2012], time budgets [@schillingerConsideringRoleTime2014], knapper skill levels [@stout2014; @herzlinger2017; @caruana2021], and mental templates [@garcía-medrano2019; @hutchence2021]. From this extensive list, knapper skill levels and mental templates have been repeatedly mentioned and discussed in the now extensive corpus of handaxe studies, and Boxgrove handaxes have been one of the most studied assemblages from these two angles. Of particular attention here are the experimental works conducted by Stout et al. [-@stout2014] focusing on inferring high knapping skill level and Garcia-Medrano et al. [-@garcía-medrano2019] identifying the mental template of the Boxgrove assemblage. Our paper combines these two perspectives and provides novel insights to the same archaeological assemblage by comparing it with experimentally made handaxes.

In its classical definition, the term mental template indicates that the "idea of the proper form of an object exists in the mind of the maker, and when this idea is expressed in tangible form in raw material, an artifact results" [@deetz1967: 45]. This concept lies at the very foundation of the cultural-historical approach in that the identification of archaeological cultures is based on the existence of distinct mental templates in a given spatial-temporal framework. Early researchers, whether explicitly or implicitly, often endorsed this conceptual framework and actively applied it in the typological analysis of handaxes at the regional level [@roe1969; @wenban-smith2000; @wenban-smith2004]. Combined with the production of large flakes, the emergence of mental templates (or "imposed form") has been recognized as a major technological innovation of the Acheulean compared with the Oldowan [@isaac1986]. For a decade or so, this concept has been less frequently used, since it was criticized for a) its normative and static assumption [@lyman2004], b) ignoring other competing factors such as raw material constraints [@white1995], and c) being constrained by the basic fracture mechanics and design space of bifacial technology [@moore2011; @moore2016]. To avoid the historical baggage associated with this controversial term, some researchers developed alternative frameworks such as "design imperatives" derived from utilitarian and ergonomic principles, which refers to a set of minimum features shared by all handaxes including their glob-butt, forward extension, support for the working edge, lateral extension, thickness adjustment, and skewness [@gowlett2006; @wynn2018]. The major difference between the concepts of design imperatives and mental templates lies in the fact that the former does not necessarily require the presence of internal representation, where handaxes can emerge "through the coalescence of ergonomic needs in the manipulation of large cutting tools [@wynn2021: 185]."

Until recently, researchers have actively addressed the above-mentioned critiques and reconceptualized the concept of mental template in the study of handaxe morphology. Regarding the normative and static assumptions, Hutchence and Scott [-@hutchence2021], for example, leveraged the theory of "community of practice" [@wenger1998] to explain the stability of Boxgrove handaxe design across multiple generations, especially how the social norms behind the consolidated material expressions were developed and negotiated by individuals in a group who have a shared history of learning. They further emphasized that emergent actions of individual knappers also contribute greatly to the shape of Boxgrove handaxes but they were simultaneously constrained by the imposition of social norms. This view also somewhat echoes the "individualized memic construct" proposed by McNabb et al. [-@mcnabb2004], which highlighted the influence of individual agency that is complementary to the traditionally favored explanation of social learning. As for the critique towards confounding factors explaining morphological variability, raw material is often treated as an important variable to be controlled at the very beginning of a research design focusing on mental templates. This is best exemplified by an experimental study of García-Medrano et al. [-@garcía-medrano2019], where they carefully chose experimental nodules mirroring those found in the Boxgrove archaeological assemblage in composition, size, and shape. Regarding the critique of design space constraint, Moore and Perston's experiment [-@moore2016] suggested that bifaces can be manufactured through flake removals dictated by a random algorithm. However, Moore [-@moore2020: 656-657] also suggested that these random experiments cannot produce "attributes like the congruent symmetries of handaxes seen in the Late Acheulean." In short, when exercised with proper caution, the concept of mental templates still has its value in our study of handaxe morphological variation, which can be further dissected into a series of shape variables corresponding to pointedness, elongation, and cross-sectional thinning among other things.

@@

The negative correlation between the PC1 and PC2 values revealed a hidden structural constraint regarding the relationship between cross-sectional thinning and the imposed form. Our results (Fig.) suggested thinner handaxes (low PC1 value) are generally more pointed/less ovate (high PC2 value). In the thinning phase of handaxe making, a knapper must strike flakes that travel more than one half way across the surface. Consequently, it would be easier to perform thinning if the plan shape of a handaxe is narrower and more pointed. It is possible that such constraints help to explain the convergence of our novice knappers on similar shapes to those preferred by modern expert knappers, however, this clearly does not explain the design target at Boxgrove. Given the ovate forms of the Boxgrove assemblage, it thus requires a high skill level to overcome this structural constraint to produce thin yet wide handaxes as demonstrated by the Boxgrove knappers. This also provides an alternative explanation to the social transmission of form for the experimental convergence of on pointed forms. In this comparative context, it would only be the Boxgrove assemblage that provided evidence of social conformity on a more difficult target shape.

In terms of our second research question, this study shows that training does have an immediate intervention effect (pre-training vs. post-training) in both PC1 (skill level) and PC2 (mental template). Nonetheless, once the training has been initiated, its effects across different assessments on both dimensions are rather non-significant. This finding provides a parallel line of evidence that corroborates what has been suggested in Pargeter et al. [-@pargeter2019] that 90 hours of training for handaxe making is still not enough for novices to reach the skill level as reflected in expert knappers, even considering the massive social support involved in the experiment set up including the direct and deliberate pedagogy and the simplified raw material procurement and preparation procedures. Methodologically speaking, this study also demonstrated that the pattern revealed by the multivariate analysis of morphometric data can nicely match with the expert knapper's 5 point grading scale of novices' knapping performances that takes multiple factors into consideration, including outcome, perceptual motor execution, and strategic understanding [See Table 2 of -@pargeter2019 for more details].

Moreover, this follow-up project further adds the samples produced by the Late Acheulean toolmaker as a new benchmark to deepen our understanding of this issue. As previously shown in Key's [-@keyHandaxeShapeVariation2019] previous finding regarding Boxgrove, it is noteworthy how constrained the range of Boxgrove assemblage morphological variation is as measured by both PC1 and PC2 even when compared with the modern expert group (**Figure** @ref(fig:GeneralPCA1)), especially given the fact that it has the largest sample size among all studied groups. Some potential explanations for this phenomenon include 1) the strong idiosyncrasy of individual expert knappers shaped by their own unique learning and practice experience; 2) the present-day skill shortage of our expert knapper as compared with Boxgrove knappers despite their multiple years of knapping practice [@milks2019]; and/or 3) modern knappers' skill level was affected by time constraints when they were requested to produce the reference collections [@schillingerConsideringRoleTime2014; @lewis2022].

@@

Although we are not the first research team to use secondary archaeological data [e.g., @keyHandaxeShapeVariation2019], we would still like to highlight here that this research project further exemplifys the potential of reusing old archaeological data in digital format to address novel research questions. In this paper, the main source of archaeological data is a collection of photos produced and curated more than 10 years ago, and the morphological variation data of the experimental collection are also derived from photographs instead of remeasurements of the original artifacts. Given the irreversible nature of archaeological excavations, digitized data, be it text, pictures, or videos, often become the sole evidence that is available for certain research questions. Yet, it has been widely acknowledged that the reuse of archaeological data has not received enough attention among researchers in our discipline [@faniel2018; @huggett2018; @moody2021]. Among many reasons preventing archaeologists from reusing published and digitized data [@sobotkova2018], the lack of a standardized practice of and motivation for data sharing is a prominent one [@marwick2018]. As stated in the method section, we addressed this issue by sharing the raw data and the code for generating the derived data on an open-access repository. Another major and legitimate concern of archaeological data reuse is their quality. In terms of this aspect, we do acknowledge the limitations of relying on photos when it comes to the more detailed technological analysis of stone artifacts, however, our paper shows that finding the appropriate research questions given the data available is key to revealing new novel insights into the studied topic. Moreover, we believe that this type of research has a strong contemporary relevance due to the continued influence of the COVID-19 on fieldwork-related travel and direct access to archaeological artifacts [@balandier2022; @ogundiran2021].

@@

Regarding the two research questions we proposed in the beginning, our case study suggested that 1) we can delineate the effects of skill level and mental template through the multivariate analysis of morphometric data, where the former is associated with cross-sectional thinning while the latter is reflected in elongation and pointedness; 2) Training has an immediate effect of convergence on shared design targets, but 90 hours of training is still not enough for novice to reach the level of expertise as reflected in modern experienced knappers, let alone the Boxgrove tool makers. At a larger theoretical level it questions the distinction between social learning of design targets vs. individual learning of the skills needed to achieve them. To illustrate, a thin cross section could be part of a mental template or design target and was explicitly instructed by our expert instructor to novices, but novices cannot fully understand nor achieve this technological goal due to the constraint of skill level, making it a robust indicator of the latter. In the future, more robust experimental studies are needed to deepen our understanding of the relationship between skill acquisition and the morphological variability of handaxes in the proper developmental context [@högberg2018] as well as their implications for the biological and cultural evolution of the hominin lineages.
